A star formation event in the leading arm of the Magellanic stream triggered by passage through the outer Galactic disk

### 2018-11-01 ms.tex
describe hierarchical inf

\newcommand{\articlename}{\textsl{Letter}}
\newcommand{\sectionname}{Section}

@@

% Shortcuts for this paper
\newcommand{\clustername}{\textsl{Price-Whelan 1}}
\newcommand{\lmcsmc}{LMC--SMC}
\newcommand{\bprp}{\ensuremath{\textrm{BP} - \textrm{RP}}}
\newcommand{\truepm}{\ensuremath{\tilde{\bs{\mu}}}}
\newcommand{\isodist}{\ensuremath{25~\kpc}}
\newcommand{\eep}{\ensuremath{\acronym{\textrm{EEP}}}}

@@

We use an ensemble Markov chain Monte Carlo (MCMC) sampler \citep[\texttt{emcee};][]{Goodman:XX, Foreman-Mackey:2013} to generate posterior samples over the parameters $(f, \bs{m}, s)$ using the likelihood defined above (Equation~\ref{eq:likelihood}), and assuming the following prior probability distributions: uniform over the domain $(-5, 5)~\masyr$ for each component of $\bs{m}$, uniform in $f$, and uniform in log-$s$ over the domain $-6 < \ln s < 4$ (with $s$ in units of \masyr).
We run the sampler with 32 walkers for 256 steps as burn-in, then reset the sampler and run for an additional 512 steps.
We then thin the resulting chains by taking every 16th sample, leaving a total of 1024 samples; We use these samples to estimate the median posterior parameter values and uncertainties, and to compute, for each star, posterior probabilities of belonging to the cluster mixture component.
For the young cluster, we find $\bs{m} = (-0.56,  0.47) \pm (0.04, 0.02)~\masyr$, $\ln s = -3.8 \pm 0.9$, and $f = 0.14 \pm 0.02$.
The bottom-left panel of the middle panels of \figurename~\ref{fig:pm-members} shows, in black (small black ellipse), the inferred median proper motion and proper-motion dispersion of the young cluster.
The proper-motion dispersion is consistent with zero, but is unmeasured; 95\% of the posterior samples have $s < 0.09~\masyr$.

@@

In detail, we start by using the \texttt{isochrones} package \citep{Morton:XXXX} to generate posterior samplings over stellar parameters for each individual source given its photometry and an interim prior.
We use the \acronym{MIST} \citep{Choi:2016} isochrone library, and \texttt{isochrones} automatically performs interpolation between the provided grid of stellar isochrones to evaluate predicted stellar parameters at a given location in the CMD.
The stellar parameters for each source are its ``equal evolutionary point'' number $\eep$ (see \acronym{MIST} documentation; \citealt{WHO}), age $\tau$, metallicity $\feh$, extinction $A_V$, and distance $D$, which can be uniquely mapped to a point in the observed CMD;
The likelihood of a given set of these parameters, $\boldsymbol{\theta} = (\eep, \tau, \feh, A_V, D)$, is then computed from the photometry and (assumed Gaussian) photometric uncertainties given the predicted photometry.

To generate posterior samples, we must also specify prior probability distributions over the parameters $\boldsymbol{\theta}$.
The priors for each parameter are summarized in \tablename~\ref{tbl:priors}.
The bounds on $\eep$ limit the isochrone to evolutionary phases between the zero-age main sequence to the terminal-age main sequence.
The prior and bounds on $A_V$ are set to prefer small and reasonable values of extinction for this moderately high Galactic latitude region.
The prior on distance assumes a uniform space density of stars.

As expected, we find that the photometry for individual sources in the lower main sequence are very poorly constrained in all parameters, and the prior tends to pull the posterior samplings to prefer closer, older stellar parameters.
This is a weakness of our methodology for performing the hierarchical inference: each source is considered independently, even though there is clear structure in the CMD (i.e. \figurename~\ref{fig:decam-cmd}).
A more correct way to do this would be to infer the stellar parameters of \emph{all} stars, the cluster hyperparameters, and the background simultaneously.
However, for the 417 stars we are using (in the red box in \figurename~\ref{fig:decam-cmd}), this model would have $\sim$2,000 free parameters, if left unmarginalized.
We are developing tools to perform star cluster parameter inference in this way (Morton et al., in prep.), but here we adopt a simple hack to allow us to instead perform individual posterior samplings and then combine into a hierarchical inference.

@@

In the end, we generate posterior samplings over the parameters $(\boldsymbol{\theta}, q)$ for each of the 417 sources in the selected region of the \decam\ CMD using \texttt{PyMultinest} \citep{TODO}, and store the value of the prior evaluated at the location of each sample.
We then use these samples and interim prior values to construct our hierarchical model.
In the hierarchical model, we assume that the stellar parameters of each primary star in the CMD region are either drawn from the cluster, or a background (stellar halo) population.
For the cluster, we assume the values are drawn from delta functions in age, metallicity, distance, and extinction, with the centroids of the delta functions $\boldsymbol{\alpha} = (\mu_\tau, \mu_{\feh}, \mu_D, \mu_{A_V})$ as hyperparameters of the hierarchical inference. \todo{Double check that we let all 4 of these free}
For the background model, we assume the same priors as specified in \tablename~\ref{tbl:priors}.
The one other parameter of this hierarchical model is the global fraction of sources that are likely cluster members, $f$.
The full XXX of this model for all $K=417$ source ... \todo{write this down}
\begin{align}
    p(\boldsymbol{\theta}_K, q_K, \boldsymbol{\alpha}, f \given \{g, i\}_K )
        &\propto
        \prod_k^K p(g_k, i_k \given \boldsymbol{\theta}_k, q_k) \,
            p(\boldsymbol{\theta}_k, q_k \given \boldsymbol{\alpha}, f) \,
            p(\boldsymbol{\alpha}) \, p(f)
\end{align}
(see \citealt{Hogg:2008, Foreman-Mackey:2013, Price-Whelan:2018} for other examples of this trick). \todo{double check references and add}
\todo{Make a PGM for this model?}

@@

We have identified a young, metal-poor stellar association in the Galactic halo --- named \clustername --- with an age $\tau \approx 150~\textrm{Myr}$, heliocentric distance $D \approx 30~\kpc$, and metallicity $\feh \approx -1.1$.
At its present day sky position, and at large distances, all significant quantities of \hi are associated with the leading arm of the Magellanic gas stream and thus it plausibly formed from Magellanic stream gas.
The age of the cluster is broadly consistent with the time it would have crossed the Galactic midplane, suggesting the possibility that interaction with the Milky Way disk or tidal compression could have triggered this star formation event.
The discovery of \clustername\ provides a critical distance constraint to the Magellanic stream and will aid future Magellanic system and Milky Way modeling efforts.
It also provides an opportunity to study star formation in a unique environment, unlike that of the Milky Way disk or any other cluster-forming region.
The serendipitous discovery of this cluster is a reminder that the combined value of the \gaia\ data with deep, large-area imaging surveys will provide many more [...].

>>>>

\newcommand{\articlename}{\textsl{Letter}}
\newcommand{\sectionname}{Section}
\newcommand{\equationname}{Equation}

@@

% Shortcuts for this paper
\newcommand{\clustername}{\textsl{Price-Whelan 1}}
\newcommand{\lmcsmc}{LMC--SMC}
\newcommand{\bprp}{\ensuremath{\textrm{BP} - \textrm{RP}}}
\newcommand{\truepm}{\ensuremath{\tilde{\bs{\mu}}}}
\newcommand{\isodist}{\ensuremath{25~\kpc}}
\newcommand{\eep}{\ensuremath{\acronym{\textrm{EEP}}}}
\newcommand{\Nisofit}{417}

@@

We use an ensemble Markov chain Monte Carlo (MCMC) sampler \citep[\texttt{emcee};][]{Goodman:XX, Foreman-Mackey:2013} to generate posterior samples over the parameters $(f, \bs{m}, s)$ using the likelihood defined above (\equationname~\ref{eq:likelihood}), and assuming the following prior probability distributions: uniform over the domain $(-5, 5)~\masyr$ for each component of $\bs{m}$, uniform in $f$, and uniform in log-$s$ over the domain $-6 < \ln s < 4$ (with $s$ in units of \masyr).
We run the sampler with 32 walkers for 256 steps as burn-in, then reset the sampler and run for an additional 512 steps.
We then thin the resulting chains by taking every 16th sample, leaving a total of 1024 samples; We use these samples to estimate the median posterior parameter values and uncertainties, and to compute, for each star, posterior probabilities of belonging to the cluster mixture component.
For the young cluster, we find $\bs{m} = (-0.56,  0.47) \pm (0.04, 0.02)~\masyr$, $\ln s = -3.8 \pm 0.9$, and $f = 0.14 \pm 0.02$.
The bottom-left panel of the middle panels of \figurename~\ref{fig:pm-members} shows, in black (small black ellipse), the inferred median proper motion and proper-motion dispersion of the young cluster.
The proper-motion dispersion is consistent with zero, but is unmeasured; 95\% of the posterior samples have $s < 0.09~\masyr$.

@@

In detail, we start by using the \texttt{isochrones} package \citep{Morton:XXXX} to generate posterior samplings over stellar parameters for each individual source given its photometry and an interim prior.
We use the \acronym{MIST} \citep{Choi:2016} isochrone library, and \texttt{isochrones} automatically performs interpolation between the provided grid of stellar isochrones to evaluate predicted stellar parameters at a given location in the CMD.
The stellar parameters for each source are its ``equal evolutionary point'' number $\eep$ (see \acronym{MIST} documentation; \citealt{WHO}), age $\tau$, metallicity $\feh$, extinction $A_V$, and distance $D$, which can be uniquely mapped to a point in the observed CMD;
The likelihood of a given set of these parameters, $\bs{\theta} = (\eep, \tau, \feh, A_V, D)$, is then computed from the photometry and (assumed Gaussian) photometric uncertainties given the predicted photometry.

To generate posterior samples, we must also specify prior probability distributions over the parameters $\bs{\theta}$.
The priors for each parameter are summarized in \tablename~\ref{tbl:priors}.
The bounds on $\eep$ limit the isochrone to evolutionary phases between the zero-age main sequence to the terminal-age main sequence.
The prior and bounds on $A_V$ are set to prefer small and reasonable values of extinction for this moderately high Galactic latitude region.
The prior on distance assumes a uniform space density of stars.

As expected, we find that the photometry for individual sources in the lower main sequence are very poorly constrained in all parameters, and the prior tends to pull the posterior samplings to prefer closer, older stellar parameters.
This is a weakness of our methodology for performing the hierarchical inference: each source is considered independently, even though there is clear structure in the CMD (i.e. \figurename~\ref{fig:decam-cmd}).
A more correct way to do this would be to infer the stellar parameters of \emph{all} stars, the cluster hyperparameters, and the background simultaneously.
However, for the \Nisofit\ stars we are using (in the red box in \figurename~\ref{fig:decam-cmd}), this model would have $\sim$2,000 free parameters, if left unmarginalized.
We are developing tools to perform star cluster parameter inference in this way (Morton et al., in prep.), but here we adopt a simple hack to allow us to instead perform individual posterior samplings and then combine into a hierarchical inference.

@@

Ultimately we generate posterior samplings over the parameters $(\bs{\theta}, q)$ for each of the \Nisofit\ sources in the selected region of the \decam\ CMD using \texttt{PyMultinest} \citep{TODO}, and store the value of the prior evaluated at the location of each sample.
We then use these samples and interim prior values to construct our hierarchical model, as described below.
In the hierarchical model, we assume that the stellar parameters of each primary star in the selected region are either drawn from the cluster, or a background (stellar halo) population.
For the cluster, we assume the values are drawn from delta functions in age, metallicity, distance, and extinction, with the centroids of the delta functions $\bs{\alpha} = (\mu_\tau, \mu_{\feh}, \mu_D, \mu_{A_V})$ as hyperparameters of the hierarchical inference.
For the background model, we assume the same priors as specified in \tablename~\ref{tbl:priors}.
The one additional parameter that must be included in this hierarchical model is the ``mixture weight'': the global fraction of sources that are likely cluster members, $f$.
To compute the likelihood of the hierarchical model, we use the individual posterior samplings to marginalize over the per-source $k$ stellar parameters $\bs{\theta}_k$ to compute the marginal likelihood $p(\bs{m}_k \given \bs{\alpha}, f)$, where $\bs{m}_k = (g, i)_k$ is the vector of photometric data;
This likelihood for a single source $k$ given a set of hyperparameters $(\bs{\alpha}, f)$ is
\begin{equation}
    p(\bs{m}_k \given \bs{\alpha}, f) = \int \dd \bs{\theta}_k \,
        p(\bs{m}_k \given \bs{\theta}_k) \,
        p(\bs{\theta}_k \given \bs{\alpha}, f) \quad .
\end{equation}
We employ the ``importance sampling trick'' (see, e.g., Appendix of \citealt{Price-Whelan:2018} or \citealt{Hogg:2008, Foreman-Mackey:2013} for other examples) to re-write an approximate form for this marginal likelihood as
\begin{equation}
    p(\bs{m}_k \given \bs{\alpha}, f) \approx \frac{\mathcal{Z}_k}{J} \,
        \sum_j^J \frac{p(\bs{\theta}_{kj} \given \bs{\alpha}, f)}{p(\bs{\theta}_{kj} \given \bs{\alpha}_0, f)}
        \label{eq:marglike}
\end{equation}
where the index $j$ specifies the index of one of $J$ posterior samples generated from the independent samplings (described above), $\mathcal{Z}_k$ is a constant, and the denominator, $p(\bs{\theta}_{kj} \given \bs{\alpha}_0, f)$, are the values of the interim prior used to do the independent samplings.
In this work, $K=\Nisofit$ and we adopt $J=2048$.

\todo{we ignore the $q$!}
With the marginal likelihood (\equationname~\ref{eq:marglike}), we then need to specify prior probability distributions for the hyperparameters $(\bs{\alpha}, f)$, and we can then generate posterior samples for the hyperparameters;
We use uniform priors for all of these, as summarized in \tablename~\ref{tbl:hyperpriors}.
We then use \texttt{emcee} \citep{Goodman:XX, Foreman-Mackey:2013} to sample from the posterior probability distribution for the hyperparameters given all of the photometric data,
\begin{equation}
    p(\bs{\alpha}, f \given \{\bs{m}_k\}) \propto
        p(\bs{\alpha}) \, p(f) \,
        \prod_k^K p(\bs{m}_k \given \bs{\alpha}, f) \quad .
\end{equation}
Here we use 64 walkers and run for an initial 128 steps to burn-in the sampler before running for a final 1024 steps.

\begin{table}[ht]
\begin{center}
    \begin{tabular}{ c | c | c }
        \toprule
        Parameter & Prior & Bounds \\
        \toprule
        $\mu_\tau$ & uniform & $(1~\textrm{Myr}, 1~\textrm{Gyr})$ \\
        $\mu_{\feh}$ & uniform & $(-2, 0)$ \\
        $\mu_{A_V}$ & uniform & $(0, 1)~\textrm{mag}$ \\
        $\mu_D$ & uniform & $(1, 100)~\kpc$ \\
        $f$ & uniform & (0, 1) \\
        \toprule
    \end{tabular}
\caption{Prior probability distributions for the hyperparameters $(\bs{\alpha}, f)$.
\label{tbl:hyperpriors}}
\end{center}
\end{table}

\todo{Make a PGM for this model?}

@@

We have identified a young, metal-poor stellar association in the Galactic halo --- named \clustername --- with an age $\tau \approx 150~\textrm{Myr}$, heliocentric distance $D \approx 30~\kpc$, and metallicity $\feh \approx -1.1$.
At its present day sky position, and at large distances, all significant quantities of \hi are associated with the leading arm of the Magellanic gas stream and thus it plausibly formed from Magellanic stream gas.
The age of the cluster is broadly consistent with the time it would have most recently crossed the Galactic midplane, suggesting the possibility that interaction with the Milky Way disk or tidal compression could have triggered this star formation event.
The discovery of \clustername\ provides a critical distance constraint to the Magellanic stream and will aid future Magellanic system and Milky Way modeling efforts.
It also provides an opportunity to study star formation in a unique environment, unlike that of the Milky Way disk or any other cluster-forming region.
The serendipitous discovery of this cluster is a reminder that the combined value of the \gaia\ data with deep, large-area imaging surveys will provide many more [...].
